---
title: "Introduction to R & Environmental Data"
subtitle: "SHM 377 — Hazardous Materials Management"
author: "Yoni Rodriguez"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-copy: true
    code-overflow: wrap
    code-fold: false
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{=html}
<style>

/* =========================================================
   Lecture Styling
   ========================================================= */

/* ---------------------------------------------------------
   Typography
   Keep lecture text consistent throughout the document
   --------------------------------------------------------- */

body {
  color: #212529;
}

blockquote {
  color: #212529;
  font-size: inherit;
  font-style: normal;
  font-weight: normal;
}

blockquote strong {
  color: #212529;
}

.text-center {
  color: #212529;
}


/* =========================================================
   Responsive Mobile Layout
   Stack ALL Quarto columns on narrow screens
   ========================================================= */

@media (max-width: 768px) {

  /* Convert every Quarto column group to a vertical layout */
  .columns {
    display: block !important;
  }

  /* Every content column becomes full width */
  .columns > .column {
    width: 100% !important;
    max-width: 100% !important;
    flex: 0 0 100% !important;
  }

  /* Remove empty spacer columns used in desktop layouts */
  .columns > .column:empty {
    display: none !important;
  }

  /* Keep images within the phone viewport */
  img {
    max-width: 100%;
    height: auto;
  }

  /* Prevent long text or captions from forcing page width */
  p,
  li,
  figcaption {
    overflow-wrap: anywhere;
  }

}

</style>
```

# Virtual Lab 01

## Introduction to R & Environmental Data

Today we will use a real environmental monitoring dataset to answer a simple question:

> **What happened to ambient particle concentrations during five days of winter monitoring?**

We will use R to:

- obtain the data
- inspect the data
- check the measurements
- calculate simple summaries
- create a visualization
- interpret what we see

The goal is not to become a programmer today.

The goal is to begin using R as a tool for environmental and hazardous materials data analysis.

# The Monitoring Question

We have repeated measurements of airborne particles collected during winter monitoring.

The measurements were collected approximately every two minutes from:

**December 27 through December 31, 2019**

Our question is:

> **What happened to ambient particle concentrations during these five days?**

Before doing any analysis, we need to understand what the dataset contains.

# The Dataset

The teaching dataset contains approximately 3,600 observations.

It contains two variables:

| Variable | Meaning |
|---|---|
| `date_time` | When the measurement was collected |
| `concentration` | Measured particle concentration |

The monitoring instrument measured particle concentration.

It did not identify the exact source of every particle.

This distinction is important.

> **Measurement does not equal source attribution.**

If concentration increases, we can observe the increase.

We cannot automatically conclude what caused it.

# Why Are We Using R?

Environmental and hazardous materials work often produces large amounts of data.

Examples include:

- air monitoring
- water sampling
- soil sampling
- surface sampling
- direct reading instrument data
- meteorological measurements
- exposure measurements

R gives us a way to organize, inspect, analyze, visualize, and communicate those data.

Another major advantage is reproducibility.

Instead of clicking through a series of menus and trying to remember what we did, we can save the commands used for the analysis.

# R and RStudio

## What is R?

R is a programming language and analytical environment used for:

- data analysis
- statistics
- visualization
- reproducible research

## What is RStudio?

RStudio is the interface we will use to work with R.

You will commonly see four areas:

### Source

Where we write and save our code.

### Console

Where R executes commands.

### Environment

Shows objects currently available in R.

### Files, Plots, Packages, and Help

Tools for navigating files, viewing graphs, managing packages, and finding documentation.

One important habit for this course is:

> **Write your commands in a script rather than relying only on the Console.**

The script becomes a record of what you did.

# Your First R Commands

We will start very simply.

Run:

```{r}
2 + 2
```

R receives the instruction and returns the result underneath the code.

Try another:

```{r}
10 / 2
```

At its most basic level, working with R means:

> We give R an instruction and R returns something.

# Creating an Object

We can also store information in R.

```{r}
x <- 5
```

Now ask R what is stored in `x`.

```{r}
x
```

The symbol:

```text
<-
```

is called the assignment operator.

You can read this:

```r
x <- 5
```

as:

> Store the value 5 in an object called `x`.

We will use this idea constantly when working with datasets.

# Getting the Dataset

Our teaching dataset is:

```text
winter-smoke-2019-12-27_to_2019-12-31.csv
```

The dataset contains measurements from December 27 through December 31, 2019.

For the first lab, we want the process of getting the dataset to be simple.

Later in the course we can expand our GitHub workflow.

# Importing the Dataset

We can use `read.csv()` to import the file.

```{r}
winter_smoke <- read.csv(
  "winter-smoke-2019-12-27_to_2019-12-31.csv"
)
```

Read that command from right to left.

R will:

1. find the CSV file
2. read the file
3. create a dataset in R
4. store the dataset under the name `winter_smoke`

We now have data inside R.

# Inspect Before You Analyze

One of the most important habits in this course will be:

> **Look at your data before analyzing it.**

We want to know:

- Did the file import correctly?
- What variables are present?
- How many observations are there?
- What does one row represent?
- Did R interpret the variables correctly?
- Do the values look reasonable?

# Look at the First Rows

Use `head()`.

```{r}
head(winter_smoke)
```

This shows the first several observations.

Look at the output.

Ask yourself:

> What does one row appear to represent?

# Look at the Last Rows

Use `tail()`.

```{r}
tail(winter_smoke)
```

Looking at both the beginning and end of a dataset can help us confirm that the file contains the time period we expected.

# Variable Names

Use `names()`.

```{r}
names(winter_smoke)
```

We expect to see:

```text
date_time
concentration
```

# Dataset Dimensions

Use `dim()`.

```{r}
dim(winter_smoke)
```

R reports:

```text
rows columns
```

The rows represent observations.

The columns represent variables.

For this dataset, each row represents one monitoring measurement.

# Examine the Structure

Use `str()`.

```{r}
str(winter_smoke)
```

This tells us several things at once, including:

- variable names
- variable types
- example values

This is one of the most useful commands to run immediately after importing a dataset.

# Did R Understand the Date and Time?

The `date_time` variable may initially be imported as text.

If that happens, R does not yet understand that those values represent time.

We can convert the variable.

```{r}
winter_smoke$date_time <- as.POSIXct(
  winter_smoke$date_time
)
```

Then check the structure again.

```{r}
str(winter_smoke)
```

This introduces another important habit:

> **Do not assume your code worked. Check.**

# Check the Measurements

Before calculating averages or making graphs, we should inspect the range of the measurements.

First, find the minimum.

```{r}
min(winter_smoke$concentration)
```

Now find the maximum.

```{r}
max(winter_smoke$concentration)
```

The observed concentration range in this teaching dataset is approximately:

```text
0.000 to 0.288
```

At this stage, we should ask:

> Do these measurements appear plausible?

# Unusual Values

Suppose one concentration is much higher than most of the others.

Should we immediately delete it?

No.

An unusual value could represent:

- a real environmental event
- a nearby source
- unusual instrument behavior
- a processing problem
- something we do not yet understand

Before removing a value, ask:

- Is it physically possible?
- Do nearby observations show a similar pattern?
- Was something happening at that time?
- Could the instrument have malfunctioned?
- Do we have enough evidence to make a decision?

> **Unusual does not automatically mean incorrect.**

# Describe the Data

Now that we have inspected and checked the dataset, we can begin summarizing it.

## Mean

```{r}
mean(winter_smoke$concentration)
```

The mean gives us one way to describe the center of the measurements.

But one number cannot show us how concentrations changed through time.

## Median

```{r}
median(winter_smoke$concentration)
```

The median gives us another measure of the center.

Compare the mean and median.

Are they similar?

If not, what might that suggest about the distribution of measurements?

## Minimum

```{r}
min(winter_smoke$concentration)
```

## Maximum

```{r}
max(winter_smoke$concentration)
```

## Summary

R can calculate several descriptive statistics at once.

```{r}
summary(winter_smoke$concentration)
```

Look at the output.

What information does R provide?

What information is still missing?

# Descriptive Statistics Have Limits

A numerical summary can tell us about:

- the center
- the range
- the distribution

But it cannot tell us when concentrations changed.

For that, we need to look at the measurements through time.

# Visualizing Concentration Through Time

Because these measurements were repeatedly collected over several days, a time series graph is a natural first visualization.

We will place:

**time on the x axis**

and

**concentration on the y axis**

```{r}
plot(
  winter_smoke$date_time,
  winter_smoke$concentration,
  type = "l",
  xlab = "Date / Time",
  ylab = "Concentration"
)
```

The graph will appear directly underneath the code.

This is one of the advantages of using a Quarto document.

The code, result, and explanation remain together.

# Read the Plot Command

Look at the code again.

```r
plot(
  winter_smoke$date_time,
  winter_smoke$concentration,
  type = "l",
  xlab = "Date / Time",
  ylab = "Concentration"
)
```

We told R:

1. use `date_time` on the x axis
2. use `concentration` on the y axis
3. connect observations with a line
4. label the x axis
5. label the y axis

# Why Does the Axis Say Only Concentration?

You may notice that we have not added a concentration unit.

That is intentional.

We still need to confirm the measurement scale and units from the original project documentation.

Until that information is verified, we should not invent a unit.

> **If we have not verified something, we should not present it as fact.**

# Look Before Explaining

Study the graph before trying to explain it.

Ask:

- Is concentration constant?
- Are there peaks?
- When do the largest peaks occur?
- Are there periods of relatively low concentration?
- Are some periods more variable?
- Do any patterns appear to repeat?

The first step is observation.

# Observation and Interpretation

A statement such as:

> Particle concentration varied throughout the monitoring period, with periods of relatively low concentration interrupted by higher measured concentrations.

is an observation based on the dataset.

A statement such as:

> Residential wood burning caused the peaks.

is a source attribution claim.

The dataset alone does not prove that.

# What Can We Conclude?

From the measurements we can examine:

- when concentrations increased
- how high measured concentrations became
- how concentrations changed through time
- whether elevated periods persisted
- how variable the measurements were

# What Can We Not Conclude?

From this dataset alone, we cannot determine:

- the exact source of every particle
- why every concentration increase occurred
- whether one source caused a particular peak

Additional information would be needed.

# What Other Data Could Help?

The monitoring results may lead us to new questions.

For example:

- What was the wind speed?
- What was the wind direction?
- What was the temperature?
- Were there atmospheric inversions?
- Were there nearby emission sources?
- Did local heating activity change?

This is how data analysis often works.

One answer produces better questions.

# Return to the Original Question

Our original question was:

> **What happened to ambient particle concentrations during five days of winter monitoring?**

We now have several ways to answer that question.

We:

- inspected the dataset
- checked its structure
- examined its measurement range
- calculated descriptive statistics
- plotted concentration through time
- interpreted the observed pattern

# The SHM 377 Data Workflow

Throughout this course, we will repeatedly use a workflow like this:

```text
Question
   ↓
Data
   ↓
Check
   ↓
Describe / Analyze
   ↓
Visualize
   ↓
Interpret
   ↓
Communicate / Manage
```

The specific commands will change.

The reasoning process will remain similar.

# Connecting Data to Hazardous Materials Management

In hazardous materials management, we may begin with questions about:

```text
Material
   ↓
Properties
   ↓
Source / Activity
   ↓
Release
   ↓
Environmental Medium
   ↓
Transport
   ↓
Exposure
   ↓
Dose
   ↓
Effects
```

Monitoring provides evidence about parts of that system.

Data analysis helps us move from measurements toward evaluation and decision making.

```text
Hazardous Materials Problem
          ↓
      Measurement
          ↓
         Data
          ↓
      Evaluation
          ↓
    Interpretation
          ↓
 Management / Control
```

# Reproducibility

One reason we are using R is that the analysis itself can be saved.

For example:

```r
winter_smoke <- read.csv(
  "winter-smoke-2019-12-27_to_2019-12-31.csv"
)

head(winter_smoke)

summary(winter_smoke$concentration)

plot(
  winter_smoke$date_time,
  winter_smoke$concentration,
  type = "l"
)
```

These commands provide a record of what we did.

Someone can inspect the code and understand how the analysis was produced.

Later, this idea will lead naturally into Quarto reports.

# You Do Not Need to Memorize Everything

Today you encountered commands such as:

```r
read.csv()
head()
tail()
names()
dim()
str()
as.POSIXct()
min()
max()
mean()
median()
summary()
plot()
```

You do not need to memorize all of these immediately.

What matters more is understanding why we use them.

Ask yourself:

```text
What is my question?

What does my dataset contain?

Did the data import correctly?

Do the measurements look reasonable?

How can I describe them?

How can I visualize them?

What does the evidence show?

What can I not conclude?
```

# What You Accomplished

In this first virtual lab, you:

- worked with a real environmental monitoring dataset
- imported data into R
- inspected the dataset
- checked the measurements
- calculated basic descriptive statistics
- created a time series visualization
- interpreted the results
- distinguished measurement from source attribution

# The Main Idea

> **We are not learning R for the sake of learning R.**

We are learning how to use data to investigate, evaluate, and communicate hazardous materials problems.

# Where We Go Next

We will continue building on the same workflow throughout SHM 377.

Over time, we will move from basic R commands toward:

```text
R
↓
R scripts
↓
data analysis
↓
visualization
↓
interpretation
↓
reproducible Quarto reports
```

We do not need to learn everything at once.

We start with the question.

Then we use the data to work toward an answer.